---
title: "Getting Started with repoquet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with repoquet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What repoquet does

repoquet turns a folder of heterogeneous source files (CSV, Stata, SPSS,
SAS, Parquet, ...) into a validated, resumable, Hive-partitioned Parquet
repository queried with DuckDB. A single workbook -- the Master Database
Table (MDT), conventionally named `DBSetup.xlsx` -- describes every source
file: which database and table it belongs to, its partition column and
value, and (optionally) a URL to download it from. The workflow itself has
seven stages, always run in the same order:

1. **Initialize** paths, configuration, and a traceable run identifier.
2. **Validate** the MDT structurally, fail fast before any repository write.
3. **Survey** source evidence and write a compact schema review workbook.
4. **Review and finalize** that workbook into an approved schema catalog.
5. **Load** the reviewed schemas into Hive-partitioned, resumable Parquet.
6. **Register** DuckDB views over the Parquet output and validate them.
7. **Audit** to reconcile the MDT, checkpoint, files on disk, and the manifest.

This vignette walks through that cycle twice: once against a small,
fully synthetic repository that runs offline in a few seconds (the part
of this vignette that actually executes), and once -- shown but not
run automatically -- against real public health datasets.

## A runnable example: the synthetic quickstart

`generate_example_repository()` builds a complete toy project in one call: a
year-partitioned `SALES` database, a site-partitioned `SENSORS` database, and
a labelled Stata `STUDY` file, plus a filled-in `DBSetup.xlsx` describing all
of them. Nothing here touches the network.

```{r}
library(repoquet)

project_dir <- tempfile("repoquet_vignette_")
invisible(generate_example_repository(project_dir))
list.files(project_dir)
```

`generate_example_repository()` returns (invisibly) the same scaffold paths
`create_repository_project()` does -- the config, the MDT, the output
locations for Parquet, checkpoints, logs, and the schema registry:

```{r}
cfg <- load_repository_config(file.path(project_dir, "repository_config.R"))
paths <- RepositoryInitialize(FormattedDBPath = cfg$FormattedDBPath)
MDT <- openxlsx::read.xlsx(cfg$MDTPath, sheet = "Sheet1")
MDT[, c("Database", "Path", "PartitionKey", "PartitionValue")]
```

### 1-2. Initialize and validate

Every stage shares one run identifier so log lines and validation records
line up across the whole run. Validation here is purely structural -- no
source file is opened yet.

```{r}
RunId <- new_repository_run_id()
ValidateMDTPreflight(MDT, strict = TRUE, ParquetBasePath = paths$ParquetBasePath,
                      MasterDBPath = cfg$MasterDBPath,
                      LogPath = paths$LogPath, RunId = RunId)
```

### 3-4. Survey, review, and finalize the schema

`PrepareSchemaRegistry()` reads a bounded sample of each source, proposes a
type for every column, and writes a review workbook. In an interactive
session you would open `SchemaReview.xlsx`'s `StartHere` sheet and resolve
any decisions it flags; the synthetic sources here are clean enough that
none are required, so `FinalizeSchemaRegistry()` can run immediately.

```{r}
invisible(PrepareSchemaRegistry(
  MDT, DBLoad = sort(unique(MDT$Database)), MasterDBPath = cfg$MasterDBPath,
  ObservationPath = paths$SchemaObservationPath,
  SchemaReviewPath = paths$SchemaReviewPath,
  n_workers = 1L, SchemaRegistryPath = paths$SchemaRegistryPath,
  LogPath = paths$LogPath, RunId = RunId))

repository_catalog <- FinalizeSchemaRegistry(
  SchemaReviewPath = paths$SchemaReviewPath,
  TableSchemaPath = paths$TableSchemaPath, strict = TRUE)
```

### 5. Load reviewed schemas into Parquet

```{r}
run_result <- ParquetBackEndCreate(
  MDT = MDT, DBLoad = sort(unique(MDT$Database)), MasterDBPath = cfg$MasterDBPath,
  completed_checkpoint = load_checkpoint(paths$CheckpointPath),
  CheckpointPath = paths$CheckpointPath, ParquetBasePath = paths$ParquetBasePath,
  LogPath = paths$LogPath, n_workers = 1L, PartitionBy = "NRows",
  RAMThreshold = 4, SAV_ROW_THRESHOLD = 1000000L, SAV_CHUNK_SIZE = 1000000L,
  SchemaRegistryPath = paths$SchemaRegistryPath, TableSchemaPath = paths$TableSchemaPath,
  ManifestPath = paths$ManifestPath, MetadataWorkbookPath = paths$ManifestWorkbookPath,
  UseSchemaCatalog = TRUE, RunId = RunId)
run_result
```

### 6. Register and query DuckDB views

```{r}
TempDirPath <- file.path(cfg$FormattedDBPath, "duckdb_temp")
dir.create(TempDirPath, recursive = TRUE, showWarnings = FALSE)
con <- open_duckdb(FormattedDBPath = cfg$FormattedDBPath, DBName = cfg$DBName,
                    TempDirPath = TempDirPath, GB = "1GB", ReadOnly = FALSE)
register_parquet_view_compile(con, ParquetBasePath = paths$ParquetBasePath,
                               tables_written = unique(repository_table_names(MDT)),
                               SchemaRegistryPath = paths$SchemaRegistryPath,
                               TableSchemaPath = paths$TableSchemaPath,
                               LogPath = paths$LogPath, RunId = RunId)

DBI::dbGetQuery(con, "SELECT year, count(*) AS n_orders FROM SALES_Orders GROUP BY year ORDER BY year")
```

### 7. Audit

`audit_repository()` non-destructively reconciles the MDT, the checkpoint,
the files actually on disk, and the manifest -- a healthy repository
reports zero issues.

```{r}
repo_audit <- audit_repository(MDT, paths$ParquetBasePath, paths$CheckpointPath,
                                paths$ManifestPath, con = con,
                                LogPath = paths$LogPath, RunId = RunId)
nrow(repo_audit$issues)
```

```{r, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
unlink(project_dir, recursive = TRUE)
```

## A richer example: real public health data (not run automatically)

The synthetic repository above is deliberately tiny so this vignette builds
in seconds with no network access. repoquet also ships a much larger,
real-data example workbook, `system.file("extdata", "DBSetup.xlsx", package =
"repoquet")`, describing:

* a demo cut of **MIMIC-III** (de-identified ICU data),
* **NHANES** Demographics, Examination, and Laboratory components across
  every cycle since 1999, and
* dozens of datasets from the **UCI Machine Learning Repository**'s health
  and medicine category.

Every row declares a `SourceURI`, `SourceLicense`, and `CitationURL` --
review those before using a dataset, and note that downloading the full
workbook can mean hundreds of files and a meaningful amount of network
traffic and disk space. It is driven by the same generic command-line
script the package ships at `system.file("scripts", "repoquet.R", package =
"repoquet")`, one subcommand per workflow stage:

```{r, eval = FALSE}
project_dir <- "my-repoquet-project"
create_repository_project(project_dir, profile = "generic")

# Point the generated repository_config.R's MDTPath at the packaged catalog,
# and MasterDBPath/FormattedDBPath at wherever you want the downloads and
# Parquet output to live, then run each stage from a shell:
```

```{sh, eval = FALSE}
Rscript system.file("scripts", "repoquet.R", package = "repoquet") validate my-repoquet-project
Rscript system.file("scripts", "repoquet.R", package = "repoquet") schema   my-repoquet-project
# Open SchemaReview.xlsx's StartHere sheet and resolve any flagged decisions, then:
Rscript system.file("scripts", "repoquet.R", package = "repoquet") finalize my-repoquet-project
Rscript system.file("scripts", "repoquet.R", package = "repoquet") load     my-repoquet-project
Rscript system.file("scripts", "repoquet.R", package = "repoquet") audit    my-repoquet-project
```

`MaterializeRemoteSources()` (called automatically by the `schema`, `load`,
and `audit` subcommands) downloads each declared source once and caches it
by content hash, so re-running any stage after the first is fast and
network-free unless a source's `DownloadPolicy` says otherwise.
